---
title: "Clustering Analyst"
author: "Chun-Han"
date: "March 14, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Clustering Analysis
The following page presents how clustering analysis will be used to identify distinct customer segments based on financial behavior, product adoption, digital engagement, and satisfaction-related variables.

## 1. Installing and loading the packages

| Library | Description |
|------------------------------------|------------------------------------|
| **pacman** | Installs and loads multiple R packages in a single command. |
| **tidyverse** | Provides tools for data wrangling, transformation, and visualization, including `dplyr`, `tidyr`, and `ggplot2`. |
| **lubridate** | Supports date parsing, date-time handling, and date arithmetic. |
| **cluster** | Provides clustering algorithms and evaluation tools, such as silhouette analysis. |
| **factoextra** | Offers helper functions for visualising clustering results, PCA outputs, and cluster quality. |
| **DT** | Creates interactive data tables for displaying cluster summaries and analytical outputs. |
| **shiny** | Provides the UI and server framework for building the interactive Shiny application. |
| **plotly** | Enables interactive visualisations for exploratory analysis and cluster-related plots. |
| **e1071** | Provides statistical utilities such as skewness computation and other descriptive measures. |
| **h2o** | Supports scalable machine learning and clustering methods, including model training on large datasets. |
| **mclust** | Implements model-based clustering, classification, and density estimation using Gaussian mixture models. |
| **parallelPlot** | Creates interactive parallel coordinate plots for comparing multivariate patterns across clusters. |

```{r}
pacman::p_load(
  pacman, tidyverse, lubridate, cluster, factoextra, DT, shiny, plotly,e1071,h2o,mclust,parallelPlot
)
```

## 2 Data Preparation

### 2.1 Loading the Dataset

We import the customer-level dataset and inspect its structure to confirm variable types.

```{r}
customers <- read_csv("data/customer_data.csv")
glimpse(customers)
```

### 2.2 Data Integrity Checks

#### 2.2.1 Missing Values

Missing values may distort clustering results. Therefore, we inspect NA counts.

```{r}
customers %>%
  summarise(across(everything(), ~sum(is.na(.))))
```

The missing value inspection shows that all 54 variables contain zero missing entries.

This indicates that the dataset is fully observed at the customer level. 
No imputation or row removal procedures are required.

#### 2.2.2 Duplicate Records

Since clustering is performed at the customer level, duplicate rows would bias the segmentation.

```{r}
sum(duplicated(customers))
```

## 3 Variable Selection for Clustering

Clustering should rely on behavioural and engagement signals, rather than identifiers or pre-existing labels.

The following variables are selected:

```{r}
cluster_vars <- c(
  "monthly_transaction_count",
  "average_transaction_value",
  "total_transaction_volume",
  "transaction_frequency",
  "avg_daily_transactions",
  "app_logins_frequency",
  "feature_usage_diversity",
  "active_products",
  "weekend_transaction_ratio",
  "customer_tenure"
)
```

## 4 Preprocessing

### 4.1 Create Original Dataset 

Before transformation, we preserve the original numeric values. This dataset will later be used to compute cluster profile summaries.

```{r}
df_original <- customers %>%
  dplyr::select(customer_id, dplyr::all_of(cluster_vars)) %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::all_of(cluster_vars),
      ~ readr::parse_number(as.character(.))
    )
  )

# Audit NA introduced by parsing
na_after_parse <- df_original %>%
  dplyr::summarise(dplyr::across(dplyr::all_of(cluster_vars), ~ sum(is.na(.)))) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "variable", values_to = "na_n") %>%
  dplyr::arrange(dplyr::desc(na_n))

na_after_parse

# Keep complete cases for clustering variables
df_original_cc <- df_original %>%
  dplyr::filter(dplyr::if_all(dplyr::all_of(cluster_vars), ~ !is.na(.)))

# Quick sanity check
tibble::tibble(
  total_rows = nrow(df_original),
  rows_used_for_clustering = nrow(df_original_cc),
  rows_removed = nrow(df_original) - nrow(df_original_cc)
)
```

### 4.2 Correlation Analysis

Before proceeding with clustering, it is crucial to check for multicollinearity among the selected variables. Highly correlated variables can distort clustering by disproportionately weighting similar information, especially in distance-based methods such as K-means and PAM, and may also reduce interpretability in model-based clustering such as GMM.

```{r}
#| code-fold: true
#| code-summary: "Code"


cor_matrix <- cor(df_original_cc %>% select(all_of(cluster_vars)),
                  use = "pairwise.complete.obs")

cor_matrix[upper.tri(cor_matrix, diag = FALSE)] <- NA

cor_long <- as.data.frame(as.table(cor_matrix)) %>%
  rename(Variable1 = Var1, Variable2 = Var2, Correlation = Freq) %>%
  drop_na()

var_order <- colnames(cor_matrix)

cor_long$Variable1 <- factor(cor_long$Variable1, levels = rev(var_order))
cor_long$Variable2 <- factor(cor_long$Variable2, levels = var_order)

ggplot(cor_long, aes(Variable1, Variable2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#7B8B97",
    mid = "#F8F9FA",
    high = "#C3A6A0",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation"
  ) +
  geom_text(aes(label = round(Correlation, 3)), color = "black", size = 3) +
  labs(
    title = "Correlation Matrix of Clustering Variables",
    x = "", y = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank()
  )
```

### 4.3 Skewness Assessment

Before applying log transformation, we quantify skewness to determine whether transformation is necessary.

```{r}
skew_table <- df_original_cc %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::all_of(cluster_vars),
      ~ e1071::skewness(., na.rm = TRUE)
    )
  ) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "variable", values_to = "skewness") %>%
  dplyr::arrange(dplyr::desc(abs(skewness)))

skew_table
```

### 4.4 Interpretation & Log Transformation

Variables with absolute `skewness > 1` are considered highly skewed and candidates for log transformation. We apply `log1p(x)` (log(1 + x)) to safely handle zero values.

```{r}
skew_vars <- skew_table %>%
  dplyr::filter(abs(skewness) > 1) %>%
  dplyr::pull(variable)

skew_vars

df_model <- df_original_cc %>%
  mutate(across(all_of(skew_vars), ~ sign(.) * log1p(abs(.))))
```

### 4.5 Skewness Comparison

To validate the effectiveness of transformation, we recompute skewness after applying log transformation.

```{r}
skew_after <- df_model %>%
  dplyr::summarise(
    dplyr::across(
      dplyr::all_of(cluster_vars),
      ~ e1071::skewness(., na.rm = TRUE)
    )
  ) %>%
  tidyr::pivot_longer(dplyr::everything(), names_to = "variable", values_to = "skew_after")

skew_compare <- skew_table %>%
  dplyr::left_join(skew_after, by = "variable") %>%
  dplyr::mutate(skew_reduction = abs(skewness) - abs(skew_after)) %>%
  dplyr::arrange(dplyr::desc(skew_reduction))

skew_compare
```

### 4.6 Standardisation

```{r}
X <- df_model %>%
  dplyr::select(dplyr::all_of(cluster_vars)) %>%
  as.matrix()

sds <- apply(X, 2, sd, na.rm = TRUE)
keep_cols <- is.finite(sds) & sds > 0
X <- X[, keep_cols, drop = FALSE]

good_rows <- apply(X, 1, function(r) all(is.finite(r)))
X <- X[good_rows, , drop = FALSE]

df_id <- df_model %>%
  dplyr::select(customer_id) %>%
  dplyr::slice(which(good_rows))

df_scaled <- scale(X)

stopifnot(all(is.finite(df_scaled)))
```

## 5 Clustering Methods

### 5.1 K-means

K-means is used as the baseline clustering method because it is computationally efficient, widely interpretable, and integrates well with Shiny-based interactive analysis.

For full-data model selection, we use H2O K-means to evaluate candidate k values on the complete standardised matrix. H2O exposes metrics such as total within-cluster sum of squares and between-cluster sum of squares for scalable K-means training.

#### 5.1.1 Full-data K selection with H2O

```{r}
# Initialise H2O
h2o::h2o.init(nthreads = -1)
h2o::h2o.no_progress()

# Convert the full scaled matrix into an H2OFrame
h2o_X <- as.h2o(as.data.frame(df_scaled))

# Candidate k values
k_grid <- 2:15

# Run full-data H2O K-means across candidate k
k_results <- purrr::map_dfr(k_grid, function(k_val) {

  km_h2o <- h2o::h2o.kmeans(
    training_frame = h2o_X,
    k = k_val,
    standardize = FALSE,   # already standardised in Section 4
    seed = 2022
  )

  tot_withinss <- as.numeric(h2o::h2o.tot_withinss(km_h2o))
  betweenss <- as.numeric(h2o::h2o.betweenss(km_h2o))
  total_ss <- tot_withinss + betweenss

  tibble(
    k = k_val,
    tot_withinss = tot_withinss,
    betweenss = betweenss,
    between_ratio = betweenss / total_ss
  )
})

k_results
```

#### 5.1.2 Model Selection Metrics: Elbow (WSS)

Before choosing k, we inspect compactness using the Elbow Method.The metric used is **Total Within-Cluster Sum of Squares (WSS)**.
A sharp reduction followed by a flattening curve suggests diminishing returns.

::: panel-tabset
### Plot

```{r}
#| echo: false

ggplot(k_results, aes(k, tot_withinss)) +
  geom_line(color = "#7B8B97", linewidth = 1) +
  geom_point(color = "#C3A6A0", size = 3) +
  labs(
    title = "Elbow Method",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_minimal()
```

### Code

```{r}
#| eval: false

ggplot(k_results, aes(k, tot_withinss)) +
  geom_line(color = "#7B8B97", linewidth = 1) +
  geom_point(color = "#C3A6A0", size = 3) +
  labs(
    title = "Elbow Method",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_minimal()
```
:::

:::callout-note
As k increases, WSS will always decrease because more clusters reduce the distance between points and their assigned centroids. Therefore, the goal is not to minimise WSS but to identify the point where the rate of improvement slows down.

The “elbow” point represents a balance between model simplicity and cluster compactness. After this point, increasing the number of clusters yields diminishing returns in terms of improved cluster cohesion.
:::

#### 5.1.3 BetweenSS / TotalSS ratio

::: panel-tabset
### Plot

```{r}
#| echo: false

ggplot(k_results, aes(k, between_ratio)) +
  geom_line(color = "#7B8B97", linewidth = 1) +
  geom_point(color = "#A8B7AB", size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "BetweenSS / TotalSS Ratio",
    x = "Number of Clusters (k)",
    y = "BetweenSS / TotalSS"
  ) +
  theme_minimal()
```

### Code

```{r}
#| eval: false

ggplot(k_results, aes(k, between_ratio)) +
  geom_line(color = "#7B8B97", linewidth = 1) +
  geom_point(color = "#A8B7AB", size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "BetweenSS / TotalSS Ratio",
    x = "Number of Clusters (k)",
    y = "BetweenSS / TotalSS"
  ) +
  theme_minimal()
```
:::

:::callout-note
This ratio measures how much of the total variance in the data is explained by the clustering structure. A higher ratio indicates that clusters are more clearly separated.

As k increases, the ratio typically increases but eventually stabilises. The optimal k is often found where the improvement becomes marginal.
:::

### 5.2 CLARA

CLARA (Clustering Large Applications) is included as a large-data medoid-based alternative to K-means.

Compared with K-means, CLARA is less sensitive to extreme values because it is based on medoids rather than means. It is also more suitable than standard PAM when the dataset is large.

For CLARA, the most natural model-selection criterion is average silhouette width.

#### 5.2.1 CLARA Model Selection

```{r}
clara_results <- purrr::map_dfr(2:15, function(k_val) {
  clara_fit <- cluster::clara(df_scaled, k = k_val, samples = 5)

  tibble(
    k = k_val,
    avg_silhouette = clara_fit$silinfo$avg.width
  )
})

clara_results
```

#### 5.2.2 CLARA Plot

::: panel-tabset
### Plot

```{r}
#| echo: false

ggplot(clara_results, aes(k, avg_silhouette)) + 
  geom_line(color = "#7B8B97", linewidth = 1) +
  geom_point(color = "#C3A6A0", size = 3) +
  labs(
    title = "CLARA Score",
    x = "Number of Clusters (k)",
    y = "Average Silhouette"
  ) +
  theme_minimal()
```

### Code

```{r}
#| eval: false

ggplot(clara_results, aes(k, avg_silhouette)) +
  geom_line(color = "#7B8B97", linewidth = 1) +
  geom_point(color = "#C3A6A0", size = 3) +
  labs(
    title = "CLARA Score",
    x = "Number of Clusters (k)",
    y = "Average Silhouette"
  ) +
  theme_minimal()
```
:::

:::callout-note
For each observation, the silhouette value ranges from −1 to 1:

- Values close to 1 indicate well-separated clusters
- Values near 0 suggest overlapping clusters
- Negative values indicate potential misclassification

The average silhouette score across all observations is computed for each k. The optimal number of clusters is typically the value of k that maximises the average silhouette score.
:::

### 5.3 GMM

Gaussian Mixture Models (GMM) are included as a model-based clustering alternative.

Unlike K-means and PAM, GMM does not assume spherical clusters and instead models the data as a mixture of Gaussian components. For GMM, model selection is based on BIC rather than elbow or silhouette.

#### 5.3.1 GMM Fitting

```{r}
gmm_fit <- mclust::Mclust(df_scaled, G = 2:15)

summary(gmm_fit)
```

#### 5.3.2 GMM BIC Plot

::: panel-tabset
### Plot

```{r}
#| echo: false
#| width: 10

par(xpd = TRUE, mar = c(5, 2, 2, 13))  

plot(gmm_fit, what = "BIC", legendArgs = list(x = "right", inset = c(-0.6, 0), cex = 0.8))

par(xpd = FALSE)
```

### Code

```{r}
#| eval: false
#| width: 10

par(xpd = TRUE, mar = c(5, 4, 4, 13))  

plot(gmm_fit, what = "BIC", legendArgs = list(x = "right", inset = c(-0.6, 0), cex = 0.8))

par(xpd = FALSE)
```
:::

## 6. K-Mean Clustering Analyst

For consistency of explanation and ease of interpretation, the downstream prototype outputs below are demonstrated using K-means with k=3.

```{r}
k_demo <- 3
method_demo <- "K-means"

set.seed(2022)
km <- kmeans(df_scaled, centers = k_demo, nstart = 50)

cluster_vector <- km$cluster

tibble(
  method_demo = method_demo,
  k_demo = k_demo,
  n_clusters = length(unique(cluster_vector))
)
```

### 6.1 Cluster Membership Output

Each customer included in the clustering workflow receives a cluster label.
To keep the downstream analysis method-agnostic, cluster assignments are stored in a generic object `cluster_vector`.

```{r}
#| code-fold: true
#| code-summary: "Code"

cluster_labels <- df_id %>%
  mutate(cluster = factor(cluster_vector))

customers_labeled <- customers %>%
  left_join(cluster_labels, by = "customer_id")

customers_clustered <- customers_labeled %>%
  filter(!is.na(cluster))

tibble(
  method_used = method_demo,
  total_customers = nrow(customers_labeled),
  clustered_customers = nrow(customers_clustered),
  excluded_customers = nrow(customers_labeled) - nrow(customers_clustered)
)
```

### 6.2 Cluster Size Distribution

Cluster size reveals whether the selected method identifies balanced segments or a smaller niche group.

```{r}
#| code-fold: true
#| code-summary: "Code"

cluster_size <- customers_clustered %>%
  dplyr::count(cluster) %>%
  dplyr::mutate(proportion = n / sum(n))

ggplot(cluster_size, aes(cluster, proportion, fill = cluster)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB", "#C3A6A0")) +
  labs(
    title = paste("Cluster Size Distribution -", method_demo),
    x = "Cluster",
    y = "Proportion of Customers"
  ) +
  theme_minimal()
```

### 6.3 Behavioural Contrast

To understand how clusters differ behaviourally, we visualise the mean standardised profile of each cluster across the selected clustering variables.

This parallel coordinates plot shows the relative behavioural shape of each cluster rather than only the magnitude of difference.

```{r}
#| code-fold: true
#| code-summary: "Code"

pp_data <- df_model %>%
  dplyr::left_join(cluster_labels, by = "customer_id") %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::select(cluster, dplyr::all_of(cluster_vars))

histoVisibility <- names(pp_data)

parallelPlot::parallelPlot(
  data = pp_data,
  refColumnDim = "cluster",
  rotateTitle = TRUE,
  histoVisibility = histoVisibility
)
```


### 6.4 Key Behavioural Drivers

To avoid subjective storytelling, we rank variables by the magnitude of difference between clusters.

```{r}
#| code-fold: true
#| code-summary: "Code"


df_profile <- df_model %>%
  dplyr::left_join(cluster_labels, by = "customer_id") %>%
  dplyr::filter(!is.na(cluster))

cluster_means_long <- df_profile %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(cluster_vars), mean),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    -cluster,
    names_to = "variable",
    values_to = "mean_value"
  )

diff_rank <- cluster_means_long %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(
    range_diff = max(mean_value) - min(mean_value),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(range_diff))

diff_rank %>% head(5)
```

### 6.5 Demographic Profiling

Segmentation becomes more meaningful when linked back to customer characteristics.

For this prototype, we examine whether cluster membership differs by gender.

```{r}
#| code-fold: true
#| code-summary: "Code"

gender_profile <- customers_clustered %>%
  dplyr::group_by(cluster, gender) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(gender_profile, aes(cluster, prop, fill = gender)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Gender Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

investment_account_profile <- customers_clustered %>%
  dplyr::group_by(cluster, investment_account) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(investment_account_profile, aes(cluster, prop, fill = investment_account)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Investment Account Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

insurance_product_profile <- customers_clustered %>%
  dplyr::group_by(cluster, insurance_product) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(insurance_product_profile, aes(cluster, prop, fill = insurance_product)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Insurance Product Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

income_bracket_profile <- customers_clustered %>%
  dplyr::group_by(cluster, income_bracket) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(income_bracket_profile, aes(cluster, prop, fill = income_bracket)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92")) +
  labs(
    title = paste("Income Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

education_level_profile <- customers_clustered %>%
  dplyr::group_by(cluster, education_level) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(education_level_profile, aes(cluster, prop, fill = education_level)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92","#E5D9D0","#8E9E94")) +
  labs(
    title = paste("Education Level Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

marital_status_profile <- customers_clustered %>%
  dplyr::group_by(cluster, marital_status) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(marital_status_profile, aes(cluster, prop, fill = marital_status)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92","#E5D9D0","#8E9E94")) +
  labs(
    title = paste("Marital Status Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

### 6.6 Distribution Comparison

To further illustrate behavioural contrast, we visualise the top-ranked variable from the driver table.

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[1]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[2]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[3]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```


## 7. CLARA Clustering Analyst

For consistency of comparison, the downstream prototype outputs below are demonstrated using CLARA with k = 3.

```{r}
k_demo <- 3
method_demo <- "CLARA"

set.seed(2022)
clara_fit <- cluster::clara(df_scaled, k = k_demo, samples = 5)

cluster_vector <- clara_fit$clustering

tibble::tibble(
  method_demo = method_demo,
  k_demo = k_demo,
  n_clusters = length(unique(cluster_vector))
)
```

### 7.1 Cluster Membership Output

Each customer included in the clustering workflow receives a cluster label.
To keep the downstream analysis method-agnostic, cluster assignments are stored in a generic object `cluster_vector`.

```{r}
#| code-fold: true
#| code-summary: "Code"

cluster_labels <- df_id %>%
  dplyr::mutate(cluster = factor(cluster_vector))

customers_labeled <- customers %>%
  dplyr::left_join(cluster_labels, by = "customer_id")

customers_clustered <- customers_labeled %>%
  dplyr::filter(!is.na(cluster))

tibble::tibble(
  method_used = method_demo,
  total_customers = nrow(customers_labeled),
  clustered_customers = nrow(customers_clustered),
  excluded_customers = nrow(customers_labeled) - nrow(customers_clustered)
)
```

### 7.2 Cluster Size Distribution

Cluster size reveals whether the selected method identifies balanced segments or a smaller niche group.

```{r}
#| code-fold: true
#| code-summary: "Code"

cluster_size <- customers_clustered %>%
  dplyr::count(cluster) %>%
  dplyr::mutate(proportion = n / sum(n))

ggplot(cluster_size, aes(cluster, proportion, fill = cluster)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Cluster Size Distribution -", method_demo),
    x = "Cluster",
    y = "Proportion of Customers"
  ) +
  theme_minimal()
```

### 7.3 Behavioural Contrast

To understand how clusters differ behaviourally, we visualise the mean standardised profile of each cluster across the selected clustering variables.

This parallel coordinates plot shows the relative behavioural shape of each cluster rather than only the magnitude of difference.

```{r}
#| code-fold: true
#| code-summary: "Code"

pp_data <- df_model %>%
  dplyr::left_join(cluster_labels, by = "customer_id") %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::select(cluster, dplyr::all_of(cluster_vars))

histoVisibility <- names(pp_data)

parallelPlot::parallelPlot(
  data = pp_data,
  refColumnDim = "cluster",
  rotateTitle = TRUE,
  histoVisibility = histoVisibility
)
```


### 7.4 Key Behavioural Drivers

To avoid subjective storytelling, we rank variables by the magnitude of difference between clusters.

```{r}
#| code-fold: true
#| code-summary: "Code"

df_profile <- df_model %>%
  dplyr::left_join(cluster_labels, by = "customer_id") %>%
  dplyr::filter(!is.na(cluster))

cluster_means_long <- df_profile %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(cluster_vars), mean),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    -cluster,
    names_to = "variable",
    values_to = "mean_value"
  )

diff_rank <- cluster_means_long %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(
    range_diff = max(mean_value) - min(mean_value),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(range_diff))

diff_rank %>% head(5)
```

### 7.5 Demographic Profiling

Segmentation becomes more meaningful when linked back to customer characteristics.

For this prototype, we examine whether cluster membership differs by gender.

```{r}
#| code-fold: true
#| code-summary: "Code"

gender_profile <- customers_clustered %>%
  dplyr::group_by(cluster, gender) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(gender_profile, aes(cluster, prop, fill = gender)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Gender Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

investment_account_profile <- customers_clustered %>%
  dplyr::group_by(cluster, investment_account) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(investment_account_profile, aes(cluster, prop, fill = investment_account)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB", "#C3A6A0", "#B18E88", "#7E8B92")) +
  labs(
    title = paste("Income Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

insurance_product_profile <- customers_clustered %>%
  dplyr::group_by(cluster, insurance_product) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(insurance_product_profile, aes(cluster, prop, fill = insurance_product)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Insurance Product Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

income_bracket_profile <- customers_clustered %>%
  dplyr::group_by(cluster, income_bracket) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(income_bracket_profile, aes(cluster, prop, fill = income_bracket)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92")) +
  labs(
    title = paste("Income Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

education_level_profile <- customers_clustered %>%
  dplyr::group_by(cluster, education_level) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(education_level_profile, aes(cluster, prop, fill = education_level)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92","#E5D9D0","#8E9E94")) +
  labs(
    title = paste("Education Level Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

marital_status_profile <- customers_clustered %>%
  dplyr::group_by(cluster, marital_status) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(marital_status_profile, aes(cluster, prop, fill = marital_status)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92","#E5D9D0","#8E9E94")) +
  labs(
    title = paste("Marital Status Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

### 7.6 Distribution Comparison

To further illustrate behavioural contrast, we visualise the top-ranked variable from the driver table.

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[1]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[2]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[3]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

## 8. GMM Clustering Analyst

For consistency of comparison, the downstream prototype outputs below are demonstrated using GMM with 3 clusters.

```{r}
method_demo <- "GMM"

gmm_fit_demo <- mclust::Mclust(df_scaled, G = 3)

cluster_vector <- gmm_fit_demo$classification
k_demo <- length(unique(cluster_vector))

tibble::tibble(
  method_demo = method_demo,
  k_demo = k_demo,
  n_clusters = length(unique(cluster_vector))
)
```

### 8.1 Cluster Membership Output

Each customer included in the clustering workflow receives a cluster label.
To keep the downstream analysis method-agnostic, cluster assignments are stored in a generic object `cluster_vector`.

```{r}
#| code-fold: true
#| code-summary: "Code"

cluster_labels <- df_id %>%
  dplyr::mutate(cluster = factor(cluster_vector))

customers_labeled <- customers %>%
  dplyr::left_join(cluster_labels, by = "customer_id")

customers_clustered <- customers_labeled %>%
  dplyr::filter(!is.na(cluster))

tibble::tibble(
  method_used = method_demo,
  total_customers = nrow(customers_labeled),
  clustered_customers = nrow(customers_clustered),
  excluded_customers = nrow(customers_labeled) - nrow(customers_clustered)
)
```

### 8.2 Cluster Size Distribution

Cluster size reveals whether the selected method identifies balanced segments or a smaller niche group.

```{r}
#| code-fold: true
#| code-summary: "Code"

cluster_size <- customers_clustered %>%
  dplyr::count(cluster) %>%
  dplyr::mutate(proportion = n / sum(n))

ggplot(cluster_size, aes(cluster, proportion, fill = cluster)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Cluster Size Distribution -", method_demo),
    x = "Cluster",
    y = "Proportion of Customers"
  ) +
  theme_minimal()
```

### 8.3 Behavioural Contrast

To understand how clusters differ behaviourally, we visualise the mean standardised profile of each cluster across the selected clustering variables.

This parallel coordinates plot shows the relative behavioural shape of each cluster rather than only the magnitude of difference.

```{r}
#| code-fold: true
#| code-summary: "Code"

pp_data <- df_model %>%
  dplyr::left_join(cluster_labels, by = "customer_id") %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::select(cluster, dplyr::all_of(cluster_vars))

histoVisibility <- names(pp_data)

parallelPlot::parallelPlot(
  data = pp_data,
  refColumnDim = "cluster",
  rotateTitle = TRUE,
  histoVisibility = histoVisibility
)
```


### 8.4 Key Behavioural Drivers

To avoid subjective storytelling, we rank variables by the magnitude of difference between clusters.

```{r}
#| code-fold: true
#| code-summary: "Code"

df_profile <- df_model %>%
  dplyr::left_join(cluster_labels, by = "customer_id") %>%
  dplyr::filter(!is.na(cluster))

cluster_means_long <- df_profile %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(cluster_vars), mean),
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    -cluster,
    names_to = "variable",
    values_to = "mean_value"
  )

diff_rank <- cluster_means_long %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(
    range_diff = max(mean_value) - min(mean_value),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(range_diff))

diff_rank %>% head(5)
```

### 8.5 Demographic Profiling

Segmentation becomes more meaningful when linked back to customer characteristics.

For this prototype, we examine whether cluster membership differs by gender.

```{r}
#| code-fold: true
#| code-summary: "Code"

gender_profile <- customers_clustered %>%
  dplyr::group_by(cluster, gender) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(gender_profile, aes(cluster, prop, fill = gender)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Gender Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

investment_account_profile <- customers_clustered %>%
  dplyr::group_by(cluster, investment_account) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(investment_account_profile, aes(cluster, prop, fill = investment_account)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB", "#C3A6A0", "#B18E88", "#7E8B92")) +
  labs(
    title = paste("Income Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

insurance_product_profile <- customers_clustered %>%
  dplyr::group_by(cluster, insurance_product) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(insurance_product_profile, aes(cluster, prop, fill = insurance_product)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB")) +
  labs(
    title = paste("Insurance Product Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

income_bracket_profile <- customers_clustered %>%
  dplyr::group_by(cluster, income_bracket) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(income_bracket_profile, aes(cluster, prop, fill = income_bracket)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92")) +
  labs(
    title = paste("Income Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

education_level_profile <- customers_clustered %>%
  dplyr::group_by(cluster, education_level) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(education_level_profile, aes(cluster, prop, fill = education_level)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92","#E5D9D0","#8E9E94")) +
  labs(
    title = paste("Education Level Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

marital_status_profile <- customers_clustered %>%
  dplyr::group_by(cluster, marital_status) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(cluster) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

ggplot(marital_status_profile, aes(cluster, prop, fill = marital_status)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#9297A0", "#D4C5B9", "#A8B7AB","#C3A6A0","#B18E88","#7E8B92","#E5D9D0","#8E9E94")) +
  labs(
    title = paste("Marital Status Bracket Composition by Cluster -", method_demo),
    x = "Cluster",
    y = "Proportion"
  ) +
  theme_minimal()
```

### 8.6 Distribution Comparison

To further illustrate behavioural contrast, we visualise the top-ranked variable from the driver table.

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[1]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[2]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

```{r}
#| code-fold: true
#| code-summary: "Code"

selected_var <- diff_rank$variable[3]

ggplot(customers_clustered,
       aes(cluster, log1p(.data[[selected_var]]), fill = cluster)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(values = c("#A8B7AB", "#C3A6A0", "#9297A0", "#D4C5B9")) +
  labs(
    title = paste("Log Distribution of", selected_var, "-", method_demo),
    x = "Cluster",
    y = "log(1 + value)"
  ) +
  theme_minimal()
```

## 9. Export Data for Shiny App

To facilitate the interactive Shiny application, we consolidate the cluster assignments from all three methods (K-means, CLARA, GMM) along with the preprocessed variables (`df_model`) and export them as a single dataset.

```{r}
shiny_base_data <- customers %>%
  dplyr::select(customer_id, gender, income_bracket, education_level, marital_status, investment_account, insurance_product,acquisition_channel, customer_segment, savings_account,credit_card, personal_loan, bill_payment_user, auto_savings_enabled) %>%
  dplyr::left_join(df_model, by = "customer_id")

readr::write_rds(shiny_base_data, "../RShiny/data/shiny_base_data.rds")
``` 

