---
title: "Take-home Exercise 2: Confirmatory Data Analysis Prototype"
author: "Cherilyn"
date: "2026-03-15"
date-modified: "last-modified"
format: html
number-sections: true
execute:
  warning: false
  message: false
---

## Overview

This prototype demonstrates one confirmatory module using raw data preparation and two hypothesis tests:

1.  H1: Regression test (behavior predictor -\> business outcome)
2.  H2: Group difference test (numeric value across categories)

## 1. Installing and Loading Required Packages

```{r}
required_packages <- c("tidyverse", "lubridate", "broom")

to_install <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(to_install) > 0) install.packages(to_install)
invisible(lapply(required_packages, library, character.only = TRUE))
```

## 2. Loading Raw Data

```{r}
customer_raw <- read_csv("data/customer_data.csv", show_col_types = FALSE)
tx_raw <- read_csv("data/transactions_data.csv", show_col_types = FALSE)
```

## 3. Data Preparation 

### 3.1 Clean transactions and engineer behavior features

```{r}
tx_clean <- tx_raw %>%
  mutate(
    date = as.Date(date),
    amount = as.numeric(amount),
    type = as.character(type),
    is_weekend = wday(date, week_start = 1) >= 6
  ) %>%
  filter(!is.na(customer_id), !is.na(date), !is.na(amount), amount > 0)

snapshot_date <- max(tx_clean$date, na.rm = TRUE)

tx_features <- tx_clean %>%
  group_by(customer_id) %>%
  summarise(
    tx_count = n(),
    total_spend = sum(amount, na.rm = TRUE),
    avg_spend = mean(amount, na.rm = TRUE),
    spend_std = sd(amount, na.rm = TRUE),
    first_tx_date = min(date, na.rm = TRUE),
    last_tx_date = max(date, na.rm = TRUE),
    active_days = n_distinct(date),
    recency_days = as.integer(snapshot_date - last_tx_date),
    tenure_days_tx = as.integer(last_tx_date - first_tx_date) + 1,
    tx_per_active_day = tx_count / pmax(active_days, 1),
    weekend_ratio = mean(is_weekend, na.rm = TRUE),
    .groups = "drop"
  )
```

### 3.2 Build transaction-type share features

```{r}
tx_type_share <- tx_clean %>%
  count(customer_id, type, name = "n_type") %>%
  group_by(customer_id) %>%
  mutate(share = n_type / sum(n_type)) %>%
  ungroup() %>%
  mutate(type_clean = paste0("share_", str_replace_all(type, "[^A-Za-z0-9]+", ""))) %>%
  select(customer_id, type_clean, share) %>%
  pivot_wider(names_from = type_clean, values_from = share, values_fill = 0)

tx_prep <- tx_features %>%
  left_join(tx_type_share, by = "customer_id")
```

### 3.3 Join with customer profile and outcomes

```{r}
customer_keep <- customer_raw %>%
  transmute(
    customer_id = as.numeric(customer_id),
    age,
    gender,
    income_bracket,
    customer_segment,
    satisfaction_score = as.numeric(satisfaction_score),
    nps_score = as.numeric(nps_score),
    churn_probability = as.numeric(churn_probability),
    customer_lifetime_value = as.numeric(customer_lifetime_value)
  )

conf_df <- customer_keep %>%
  inner_join(tx_prep, by = "customer_id")

dim(conf_df)
```

## 4. Variable Setup

```{r}
ALPHA <- 0.05
exclude_cols <- c("customer_id", "first_tx_date", "last_tx_date")
analysis_cols <- setdiff(names(conf_df), exclude_cols)

numeric_cols <- analysis_cols[vapply(conf_df[analysis_cols], is.numeric, logical(1))]
categorical_cols <- analysis_cols[!vapply(conf_df[analysis_cols], is.numeric, logical(1))]

pick_col <- function(preferred, pool) {
  hit <- preferred[preferred %in% pool]
  if (length(hit) > 0) hit[1] else pool[1]
}

fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "N/A", formatC(x, digits = digits, format = "f"))
}
```

## 5. H1: Regression (Behavior -\> Outcome)

Hypotheses: - H0: Slope of predictor is zero when predicting outcome. - H1: Slope of predictor is not zero.

Justification: Linear regression is suitable because both variables are numeric and the objective is to estimate direction, magnitude, and significance of association.

### 5.1 H1 function and run

```{r}
run_h1_regression <- function(data, outcome_var, predictor_var, alpha = 0.05) {
  dff <- data %>%
    transmute(
      outcome = as.numeric(.data[[outcome_var]]),
      predictor = as.numeric(.data[[predictor_var]])
    ) %>%
    drop_na()

  stopifnot(nrow(dff) >= 10, length(unique(dff$predictor)) >= 2)

  fit <- lm(outcome ~ predictor, data = dff)
  fit_tidy <- broom::tidy(fit)
  fit_glance <- broom::glance(fit)

  slope <- fit_tidy %>% filter(term == "predictor") %>% pull(estimate)
  p_value <- fit_tidy %>% filter(term == "predictor") %>% pull(p.value)
  r_squared <- fit_glance$r.squared
  decision <- ifelse(p_value < alpha, "Reject H0", "Fail to reject H0")

  list(
    data = dff,
    model = fit,
    slope = slope,
    p_value = p_value,
    r_squared = r_squared,
    decision = decision
  )
}

h1_outcome <- pick_col(c("customer_lifetime_value", "churn_probability", "total_spend"), numeric_cols)
h1_predictor <- pick_col(c("total_spend", "tx_count", "recency_days", "avg_spend"), setdiff(numeric_cols, h1_outcome))

h1_res <- run_h1_regression(conf_df, h1_outcome, h1_predictor, alpha = ALPHA)

tibble(
  Metric = c("Outcome", "Predictor", "Slope", "p-value", "R-squared", "Decision"),
  Value = c(
    h1_outcome, h1_predictor,
    fmt_num(h1_res$slope, 4),
    fmt_num(h1_res$p_value, 4),
    fmt_num(h1_res$r_squared, 3),
    h1_res$decision
  )
)
```

### 5.2 H1 plot (compressed for report size)

```{r}
set.seed(608)

plot_n <- min(5000, nrow(h1_res$data))
plot_df <- dplyr::slice_sample(h1_res$data, n = plot_n)

ggplot(plot_df, aes(x = predictor, y = outcome)) +
  geom_point(alpha = 0.35, color = "#0A9396", size = 0.9) +
  geom_smooth(method = "lm", se = FALSE, color = "#9B2226", linewidth = 1) +
  theme_minimal(base_size = 11) +
  labs(
    title = paste(h1_outcome, "vs", h1_predictor),
    x = h1_predictor,
    y = h1_outcome
  )
```

Interpretation note: If p-value \< 0.05, there is evidence that the selected behavior predictor is associated with the selected outcome.

## 6. H2: Group Difference (Value Across Groups)

Hypotheses:

-   H0: Mean value is equal across groups.

-   H1: At least one group differs.

Justification: Use ANOVA when assumptions hold; otherwise use Kruskal-Wallis for robust non-parametric comparison.

### 6.1 H2 function and run

```{r}
run_h2_group_difference <- function(data, group_var, value_var, alpha = 0.05, min_group_n = 3) {
  dff <- data %>%
    transmute(
      group = as.character(.data[[group_var]]),
      value = as.numeric(.data[[value_var]])
    ) %>%
    drop_na()

  counts <- table(dff$group)
  keep <- names(counts[counts >= min_group_n])
  dff <- dff %>% filter(group %in% keep)

  split_vals <- split(dff$value, dff$group)
  stopifnot(length(split_vals) >= 2)

  normality_pass <- TRUE
  if (min(lengths(split_vals)) < 20) {
    set.seed(42)
    for (vals in split_vals) {
      sample_vals <- if (length(vals) > 500) sample(vals, 500) else vals
      if (length(unique(sample_vals)) < 3 || shapiro.test(sample_vals)$p.value < alpha) {
        normality_pass <- FALSE
        break
      }
    }
  }

  var_p <- bartlett.test(value ~ group, data = dff)$p.value
  variance_pass <- var_p >= alpha

  if (normality_pass && variance_pass) {
    test_name <- "One-way ANOVA"
    fit <- aov(value ~ group, data = dff)
    stat <- summary(fit)[[1]]["group", "F value"]
    p_value <- summary(fit)[[1]]["group", "Pr(>F)"]
  } else {
    test_name <- "Kruskal-Wallis"
    kw <- kruskal.test(value ~ group, data = dff)
    stat <- as.numeric(kw$statistic)
    p_value <- kw$p.value
  }

  decision <- ifelse(p_value < alpha, "Reject H0", "Fail to reject H0")

  p <- ggplot(dff, aes(x = group, y = value, color = group)) +
    geom_boxplot(outlier.alpha = 0.45, show.legend = FALSE) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1)) +
    labs(
      title = paste(value_var, "by", group_var),
      x = group_var,
      y = value_var
    )

  list(
    data = dff,
    test_name = test_name,
    statistic = stat,
    p_value = p_value,
    decision = decision,
    plot = p
  )
}

h2_group <- pick_col(c("customer_segment", "income_bracket", "gender"), categorical_cols)
h2_value <- pick_col(c("total_spend", "customer_lifetime_value", "tx_count"), numeric_cols)

h2_res <- run_h2_group_difference(conf_df, h2_group, h2_value, alpha = ALPHA)

tibble(
  Metric = c("Group variable", "Value variable", "Test used", "Statistic", "p-value", "Decision"),
  Value = c(
    h2_group, h2_value,
    h2_res$test_name,
    fmt_num(h2_res$statistic, 3),
    fmt_num(h2_res$p_value, 4),
    h2_res$decision
  )
)
```

### 6.2 H2 plot

```{r, fig.width=7, fig.height=4, dpi=120}
h2_res$plot
```

Interpretation note: If p-value \< 0.05, at least one group differs in the selected metric.
